---
title: "Lab 4"
author: "Ethan Wanlass"
date: "10/24/2019"
output: html_document
---

## Background
An excerpt of the individual player data of the Cyclone football team is available from Canvas. Dowload the [excel file](cyclonesFootball2018.xlsx) and put the file next to your Rmd files. 

Some abbreviations for the column names can be found in the Glossaries sheet in the Excel file.

**References:**

- [Iowa State University Athelete](https://cyclones.com/index.aspx)
- [A Beginner's Guide to American Football](https://www.youtube.com/watch?v=3t6hM5tRlfA) for the rules of American football.

## Loading Data: Cyclone Football 

The Cyclone football data involves multiple Excel sheets. The following loads each sheet into a dataframe as well as load the libraries that I will be using. 

```{r, results = 'hide'}
library(readxl)
library(dplyr)
library(tidyr)

defense <- read_excel('cyclonesFootball2018.xlsx', sheet='Defensive')
offense <- read_excel('cyclonesFootball2018.xlsx', sheet='Offensive')
bio <- read_excel('cyclonesFootball2018.xlsx', sheet='Biography')
```

## Cleaning data

The following code will give us a glimpse of what each of the dataframes look like. 

```{r}
head(defense)
head(offense)
head(bio)
```

The first step to being able to analyze this data is to clean it. To start I factored the names in the dataframes defense, offense and bio. I also factored the opposing team's name in defense and offense. 

```{r}
defClean = defense
offClean = offense
bioClean = bio

defClean$Name = factor(defClean$Name)
offClean$Name = factor(offClean$Name)
bioClean$Name = factor(bioClean$Name)

defClean$Opponent_Opponent = factor(defClean$Opponent_Opponent)
offClean$Opponent_Opponent = factor(offClean$Opponent_Opponent)
```

Afterwards, I changed the rest of the variables in defense and offense as well as weight as a numeric variable instead of a character variable. Then did a quick check to make sure these columns were now the correct class.   

```{r}

defClean = defClean %>% mutate_at(3:11, as.numeric, na.rm = TRUE)
offClean = offClean %>% mutate_at(3:12, as.numeric, na.rm = TRUE)
bioClean = bioClean %>% mutate_at('Weight', as.numeric, na.rm = TRUE)

class(defClean$Tackles_Solo)
class(offClean$Rushing_ATT)
class(bioClean$Weight)

```

In order to change the height into a better format, I seperated it into two columns: feet and inches. 

```{r}
bioClean = bioClean %>% 
  separate(Height, into = c('Feet', 'Inches'), sep = "-")
bioClean = bioClean %>% mutate_at(c('Feet','Inches'), as.numeric, na.rm = TRUE)
class(bioClean$Feet)
```

Finally, I checked to see the structure of each of the cleaned data sets. 

```{r}
str(defClean)
str(offClean)
str(bioClean)
```

## Tidying Defense

First thing I did was tidy the columns into statistic types and the amount each player made. 

```{r}
defClean = defClean %>% gather(key=Statistic, value=Amount, 3:11)
defClean
```

The following will display a histogram of each defensive statistic:

```{r}
library(ggplot2)
ggplot(defClean, aes(Amount)) + facet_wrap(~Statistic) + geom_histogram(binwidth = .5) 
```

Looking at these histograms we can see that Sacks, QB hurries, Passes blocked and Turnovers are rarer defensive stats. 

Next, let's show the defensive performance between two specific games: West Virginia and Kansas State. We will base defensive performance on the number of solo tackles by each player. The larger number of solo tackes means a better performance. 

```{r}
solo_tackles = defClean %>% 
  filter(Statistic == 'Tackles_Solo') 
spreaded = solo_tackles %>% spread(key=Opponent_Opponent, value=Amount)
class(spreaded$`West Virginia`)
colnames(spreaded)[colnames(spreaded)=="West Virginia"] <- "West_Virginia"
colnames(spreaded)[colnames(spreaded)=="Kansas State"] <- "Kansas_State"
ggplot(spreaded, aes(West_Virginia,Kansas_State), na.rm = TRUE) + geom_point() + labs(title = 'Number of Solo tackles in the West Virginia and Kansas State game', x = 'West Virginia', y = 'Kansas State')
```

Looking at the scatterplot, there appears to be have been a slightly better defensive performance in the Kansas State game. 


## Tidying Bio (B)
Now let's turn our attention to the player biography `bioClean`.

Let's seperate the column hometown into 2 columns: city and State

```{r}
bioClean = bioClean %>% 
  separate(Hometown, into = c('City', 'State'), sep = ",")
bioClean
```

From there we can check  the count of total players from each state. 

```{r}
bioClean %>% group_by(State) %>% summarize(Number=n())
```

From our table we can see that Iowa is where the clear majority are from followed by Ohio and Texas respectively. 

## Joining DataFrames

Let's take a look at where each offensive player is from. To do so we combined the offense and bio data frames into a new dataframe and then grouped by player to see their hometowns. 
```{r}
joint = left_join(offense,bio,by = "Name")
joint%>% group_by(Name) %>% summarize(hometown = unique(Hometown))
```

Let's check if there is a relationship between 
```{r}
joint$Weight = factor(joint$Weight) 
joint$Receiving_YDS = as.numeric(joint$Receiving_YDS) 

str(joint)
ggplot(joint, aes(Weight,Receiving_YDS)) + geom_boxplot() + labs(title = 'Number of receiving yards for each Weight', x = 'Weight', y = 'Receiving Yards')
```

Looking at these boxplots, weight does not seem to have much impact on the amount of receiving yards. The large differences are probably from individual players.  

3. Explore more. Raise a question that can be answered only by looking at two or more data frames. Then answer that question.

Is there a relationship between the class and the number of touchdown receptions? 
```{r}
joint$Class = factor(joint$Class, levels = c('Freshman', 'Redshirt Freshman', 'Sophmore', 'Redshirt Sophmore', 'Junior', 'Redshirt Junior', 'Senior', 'Redshirt Senior'))

joint$Receiving_TD = as.numeric(joint$Receiving_TD)
ggplot(joint, aes(Class, Receiving_TD)) + geom_boxplot() + coord_flip() + labs(title = 'Number of receiving touchdowns per Class', x = 'Receiving Touchdowns', y = 'Class')
```

Looking at this boxplot we can see that outside of Redshirt juniors, there were very little touchdown receptions from the Iowa State Offense. 
